

---
title: "The Costs of Healthcare"
author: "Jake Johnson, Travis Barrick, Erica Fitzgerald, Carrie Baggett"
date: "December 11, 2016 [Use keyboard arrows to navigate presentation]"
output: 
  ioslides_presentation:
    fig_width: 7
    fig_height: 9
    widescreen: true
    smaller: true

runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE)
```

```{r, echo=FALSE,message=FALSE, fig.height=5}
#Libraries
library(readr)
if (!require(shiny)) install.packages("shiny")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(dplyr)) install.packages("dplyr")
if (!require(plotly)) install.packages("plotly")
if (!require(leaflet)) install.packages("leaflet")
if (!require(htmltools)) install.packages("htmltools")
if (!require(DT)) install.packages("DT")
if (!require(stringr)) install.packages("stringr")
if (!require(ggmap)) install.packages("ggmap")
if (!require(devtools)) install.packages("devtools")
if (!require(ggthemes)) install.packages("ggthemes")
library(shiny)
library(ggplot2)
library(dplyr)
library(leaflet)
library(htmltools)
library(DT)
library(stringr)
library(ggmap)
library(devtools)
library(ShinyRatingInput)
library(ggthemes)
library(plotly)

##### Data Import #####
mapstates <- read_csv("mapstates3.csv")
mapstates2<-read_csv("mapstates4.csv")
Analysis_Dataset <- read_csv("Analysis_Dataset2.csv")
AverageCategory <- read_csv("AverageCategory.csv")
states<-map_data("state")

```
##Introduction
Healthcare affects us all.  When we or a loved one face a health issue we strive to find the best care possible.  Unfortunately, US health care costs are among the highest in the world and often based on where we are located and our health affliction choices can be limited and vary based on cost.  Expenses not covered by insurance can put patients in tough financial situations.  Hospitals have the right to charge as they see fit for medical procedures and it’s difficult for patients to know if they are being billed fairly.  It is with this notion of mysterious health care costs that we explored the ‘Hospital Charge Data’ from data.gov [link](https://www.data.gov/health/).  This data exploration provides a greater understanding of the relationship between common healthcare procedures and their associated costs, availability, and quality across the United States and how healthcare costs across the nation compare to Iowa.

##Data

The ‘Hospital Charge Data’ contains billing data on the 100 most common inpatient and 30 most common outpatient services for hospitals and medical centers across the US in 2011. The services were consolidated into the 14 following procedure categories for the analysis:   Ears/Nose/Throat, Endocrine, Gastrointestinal, Heart, Infections, Lung, Nerve, Orthopedic, Other, Poisoning, Psychiatric, Renal, Skin, and Vascular.    

The dataset includes three different billing charges used in this analysis:  

- **Average Insurance Cost:**  the average amount covered by insurance  

- **Average Medicare Coverage:** the average Medicare payment for a procedure, deemed the reasonable charge as Medicare has rules they apply when determining how much will be paid for procedures  

- **Average Patient Owed:** this is the average amount owed by patients after insurance and Medicare is applied.  

To assess hospital quality, a 1 through 5 ranking was applied to the dataset using hospital rank from the “Hospital Compare” dataset provided by data.medicare.gov [link](https://data.medicare.gov/data/hospital-compare ).  Lastly, cost of living adjustments were applied to the dataset using data from the US Department of Labor to determine cost variation in relation to Iowa.  

- Access to our data and code is available on Google Drives through this [link](https://drive.google.com/drive/folders/0B8tc7fB6TuONR29hVDVaNXJEX1k?usp=sharing). *Note: You must be logged in to your Iowa State account in order to view.* 

##Our Findings
Starting broadly, we found that across the United States infections were on average the most costly procedure to treat followed by orthopedic, nerve, and gastrointestinal.  Hospitals in states such as California and Arizona charged the most on average to treat infections.  Treatment availability and quality vary greatly based on procedure, but following our infection example 5 star quality treatment options were available at 57 hospitals across the United States; costs for treatments at these hospitals range from $24k to $225k.  The cost of treating infections in Iowa is at the low end of that range, with an average cost of treatment for infections of under $32k at Mercy Hospital in Iowa City. California tops the other end of the range, with the only two hospitals with charges exceeding $200k. 

##Your Turn
This application provides four different charts to interact with and explore the health-care data:  

-**Average Cost of Healthcare by State:**  Interested in knowing differences in costs for heart procedures across states in the US?  This multi-dimensional scatterplot allows you to view the 3 primary billing components side-by-side per state by procedure category.   

-**Treatment Cost:** Using this chart you can select a procedure category and the map will populate the location of hospitals that treat it; click on the *Most Affordable Hospital* tab to see which hospitals offer the lowest cost for each procedure.  Another feature of this chart is the ability to select a hospital quality ranking of 1 to 5 stars.  

-**Cost of Healthcare Compared to Iowa:**  This map allows you to see the relationship between procedure cost categories in comparison to Iowa, both on a straight cost and cost of living adjusted basis.  

-**Number of Patients by Category in Iowa:** This scatterplot provides an additional level of detail displaying all procedures within a specified procedure category by the number of patients.  


##Average Cost of Healthcare|By State

```{r, echo=FALSE,message=FALSE,fig_width=13}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      selectInput("Category", "Select Category", 
                  choices=c("Ears,Nose,Throat",	"Endocrine",	"Gastrointestinal",	"Heart",	"Infections",	"Lung",	"Nerve",	"Orthopedic",	"Other",	"Poisoning",	"Psychiatric",	"Renal",	"Skin",	"Vascular"
                  ),
                  selected="Infections"
                  ,
                  multiple=FALSE), 
    
      width=3
    ),
    mainPanel(
      plotlyOutput("plot",width="100%"),
       p("This chart displays average costs for healthcare procedures by state.  Each point in the multi-dimensional scatterplot represents a state where a specific procedure took place.  By hovering over the points you can see and compare 3 average costs: Average Insurance Charge, Average Medicare Coverage, and Average Patient Owed Costs. Through the exploration we found that on average hospitals were charging the most to treat infections, with average cost highest in California, but a relatively low cost for patients only owing under $1500 on average. Conversely, Hawaii hospitals charged less for treatmenting infections, but more of the financial burden was passed onto the patient, with over $3k owed on average.")
    )
  )
)

server <- function(input, output,session) {
  catreact<-reactive({input$Category})
  
  
  output$plot <- renderPlotly({
    m<-Analysis_Dataset%>%
      filter(Category==catreact())%>%
      group_by(Category,ProviderState) %>% 
      summarise(avg_insurance = paste("$",prettyNum(mean(AverageCoveredCharges),big.mark=",",scientific=FALSE), sep=""),
                avg_medicare = paste("$",prettyNum(mean(AverageMedicarePayments),big.mark=",",scientific=FALSE), sep="") ,
                avg_patient_owed = paste("$",prettyNum(mean(AverageTotalPayments-AverageMedicarePayments),big.mark=",",scientific=FALSE), sep="") )
    
    p <- plot_ly(m, x = ~avg_insurance, y = ~avg_medicare, 
                 z = ~avg_patient_owed,hoverinfo = 'text',text=~paste('Category: ',Category,'</br> State: ',ProviderState ,'</br> Average Insurance Cost: ',avg_insurance,'</br> Average Medicare Coverage: ',avg_medicare,'</br> Average Patient Owed: ',avg_patient_owed),color = ~Category 
    )%>%
      add_markers(marker = list(size = 10,
                       color = 'red',
                       line = list(color = 'dark red',
                                   width = 2))) %>%
      layout(scene = list(xaxis = list(title = 'Average Insurance'),
                          yaxis = list(title = 'Average Medicare'),
                          zaxis = list(title = 'Average Patient Owed')))
    
  })
}

shinyApp(ui = ui, server = server)
```

##Treatment Cost

```{r, echo=FALSE,message=FALSE}
ui <- fluidPage(
  # Application title
  sidebarLayout(
    sidebarPanel(
      radioButtons("Category", "Select Category", 
                  choices=c("Ears,Nose,Throat",	"Endocrine",	"Gastrointestinal",	"Heart",	"Infections",	"Lung",	"Nerve",	"Orthopedic",	"Other",	"Poisoning",	"Psychiatric",	"Renal",	"Skin",	"Vascular"
                  ),
                  selected="Heart" 
                  ), 
      ratingInput("rating1", label="Min Hospital Rating",value=1),
      width=4
       ),
    
    # Show a plot of the generated distribution
    mainPanel(tabsetPanel(
      tabPanel("Map",
               leafletOutput('myMap')),
      tabPanel("Most Affordable Hospitals",
               DT::dataTableOutput("table"))),
      p("Common health procedures vary in cost by location and hospital.  The chart above provides 2 views of treatment costs.  The first is an interactive map of the United States narrowed by treatment category.  By clicking on a specific number or point, you're able to drill down to specific hospitals gleaning more information including name, address, city, average treatment cost, national treatment cost rank, and overall hospital rating. Clicking over to the Most Affordable Hospitals tab, you'll see listing of hospitals starting with the lowest average cost per treatment.  Using the filters on the left navigation you can choose different procedure categories and select 1 to 5 star ratings. Expanding upon the infection example, using this chart we can see there are 58 hospitals in the US with 5 star treatment ratings performing infection treatments; the most affordable hospital in this subset is charging $24k on average.")
    )
  )
)




server <- function(input, output,session) {
  catreact<-reactive({input$Category})
  ratingreact<-reactive({input$rating1})
  
  output$myMap <- renderLeaflet(
    test<-AverageCategory%>%
      filter(Category==catreact(), Rating >= ratingreact())%>%
      group_by(Category)%>%
      mutate(rank=row_number(),CostofTreatment=sprintf("$%s",format(AverageCost,big.mark=",",digits=2,scientific=FALSE)))%>%
      mutate(X=paste0('<strong>Provider Name: </strong>', 
                      ProviderName,
                      '<br><strong>Address</strong>: ',
                      ProviderStreetAddress,
                      '<br><strong>City</strong>: ',
                      ProviderCity,
                      '<br><strong>Average Treatment Cost</strong>: ',
                      CostofTreatment,
                      '<br><strong>National Treatment Cost Rank</strong>: ',
                      rank,
                      '<br><strong>Overall Hospital Rating</strong>: ',
                      Rating
      ))%>%
      leaflet()%>%
      addProviderTiles("Esri.NatGeoWorldMap")%>%
      setView(-110, 45, zoom = 3)%>%
      addMarkers(lng=~longitude, lat=~latitude,popup=~as.character(X),clusterOptions = markerClusterOptions())%>%
      mapOptions(zoomToLimits = "first")
  )
  
  output$table <- DT::renderDataTable(DT::datatable({
    data <-     AverageCategory%>%
      filter(Category==catreact(), Rating >= ratingreact())%>%
      mutate(CostofTreatment=sprintf("$%s",format(AverageCost,big.mark=",",digits=2,scientific=FALSE)),State=Name)
    
    
    data[,c("ProviderName","ProviderCity","State","CostofTreatment","Rating")]
    
    
  }))
  output$rating <- renderText({paste(input$rating) })

}

# Run the application
shinyApp(ui = ui, server = server)
```

##Cost of Healthcare Compared to Iowa

```{r, echo=FALSE,message=FALSE}
ui <- fluidPage(
  #titlePanel("Healthcare Expenses Compared to Iowa by Category"),
  
  sidebarLayout(
    sidebarPanel(
      radioButtons("Category", "Select Category", 
                  choices=c("Ears,Nose,Throat",	"Endocrine",	"Gastrointestinal",	"Heart",	"Infections",	"Lung",	"Nerve",	"Orthopedic",	"Other",	"Poisoning",	"Psychiatric",	"Renal",	"Skin",	"Vascular"
                  ),
                  selected="Heart"
                  ),width=3
    ),
    mainPanel(tabsetPanel(
      tabPanel("% Difference to Iowa",
               plotlyOutput("plot2",width=800,height=400
               )),
      tabPanel("Cost of Living Adjusted % Difference to Iowa",
               plotlyOutput("plot",width=800,height=400
               ))),
      p("These two charts demonstrate a cost comparison relative to the state of Iowa. The first chart shows the cost difference on a percentage basis; a 10% difference indicates that, on average, that set of procedures would be 1.1x as expensive as the same procedure in Iowa. While it's great to know proportional difference in procedural costs, the analysis is further enhanced by reviewing a Cost of Living Adjusted (COLA) proportional difference, as demonstrated with the midpoint-difference map on the second tab.


As an example of how this tool works, select the Heart category on the Cost of Living Adjusted % Difference to Iowa tab, then hover over South Dakota. This analysis shows that South Dakota heart procedures, on average, are 10.4% more expensive than the same procedures in Iowa, adjusted for cost of living in both areas. 
")
    )
  )
)  


server <- function(input, output,session) {
output$plot <- renderPlotly({
    states<-map_data("state")
    
    
    catreact<-reactive({input$Category})
    
    mapstates %>% 
      filter(Category==catreact())%>%
      group_by(Category,ProviderState,Mean_Cost_Percentage_Difference_to_Iowa)%>%
      rename(Compared.to.Iowa=Mean_Cost_Percentage_Difference_to_Iowa)%>%
      ggplot(aes(x = long, y = lat,group=group)) +   
      geom_path()+
      geom_polygon(aes(group = group, fill=Compared.to.Iowa))+
      expand_limits(x=mapstates$long, y=mapstates$lat)+
      scale_fill_gradient2(low="dark green",mid="white",high="dark red",name="Cost of Living \n% Compared to Iowa")+
      theme(plot.title = element_text(hjust = -0.7))+
      ggthemes::theme_map()
    
  })
  output$plot2 <- renderPlotly({
  
    
    catreact<-reactive({input$Category})
    mapstates2 %>% 
      filter(Category==catreact())%>%
      group_by(Category,ProviderState,Mean_Cost_Percentage_Difference_to_Iowa)%>%
      rename(Compared.to.Iowa=Mean_Cost_Percentage_Difference_to_Iowa)%>%
      ggplot(aes(x = long, y = lat,group=group)) +   
      geom_path()+
      geom_polygon(aes(group = group, fill=Compared.to.Iowa))+
      expand_limits(x=states$long, y=states$lat)+
      scale_fill_gradient2(low="dark green",mid="white",high="dark red",name="% Compared to Iowa")+
      theme(plot.title = element_text(hjust = -0.7))+
      ggthemes::theme_map()
    
  })
}

shinyApp(ui = ui, server = server)
```

##Number of Patients by Category


```{r, echo=FALSE,message=FALSE}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      selectInput("State", "Select State", 
                  choices=c("AK"	,"AL"	,"AR"	,"AZ"	,"CA"	,"CO"	,"CT"	,"DC"	,"DE"	,"FL"	,"GA"	,"HI"	,"IA"	,"ID"	,"IL"	,"IN"	,"KS"	,"KY"	,"LA"	,"MA"	,"MD"	,"ME"	,"MI"	,"MN"	,"MO"	,"MS"	,"MT"	,"NC"	,"ND"	,"NE"	,"NH"	,"NJ"	,"NM"	,"NV"	,"NY"	,"OH"	,"OK"	,"OR"	,"PA"	,"RI"	,"SC"	,"SD"	,"TN"	,"TX"	,"UT"	,"VA"	,"VT"	,"WA"	,"WI"	,"WV"	,"WY"
                  ),
                  selected="IA", 
                  multiple=FALSE),
       selectInput("Category", "Select Categories", 
                  choices=c("Ears,Nose,Throat",	"Endocrine",	"Gastrointestinal",	"Heart",	"Infections",	"Lung",	"Nerve",	"Orthopedic",	"Other",	"Poisoning",	"Psychiatric",	"Renal",	"Skin",	"Vascular"
                  ),
                  selected="Heart",
                  multiple=TRUE
                  ),width=2
    )
  ,
    mainPanel(
      p("Hover over points for information"),
        plotlyOutput("scatterplot"),
     p("For patients seeking care for a given category of ailment, such as heart issues, lung concerns, or even poisoning, one may be concerned about the amount of experience a given care facility may have in the area of need. To examine the experience data for our Iowa hospitals, the number of procedures by category can be selected and even compared using this dot plot. 


For the data used in this analysis, the hospital in Iowa most experienced with treating infections is Mercy Medical Center in Sioux City, having seen 340 patients for septicemia or severe sepsis."),
      width = 10
    )
)
)
server <- function(input, output,session) {

  
  output$scatterplot <- renderPlotly({
    m<-Analysis_Dataset %>%
      filter(ProviderState==input$State,Category==input$Category)%>%
      group_by(ProviderState,ProviderName,Category,Procedure) %>%
      summarize(NumberofPatients = sum(TotalDischarges))
    
    gg<-ggplotly(ggplot(m,aes(Procedure,NumberofPatients,label=ProviderName))+
      geom_point(aes(color=Category)) + 
      ylab("# of Patients")+
      coord_flip()+
      guides(color=FALSE))
gg
  })
  
  

}
shinyApp(ui = ui, server = server)

```

